Clustering Neural Populations by Poisson Dynamic Factor Model

### 2022-01-18 latexDraft/draft.tex
add appendix

@@

	\subsection{Poisson Dynmic Factor Model}
	Denote the observed spike-count of neuron $i \in \{ 1,\ldots,N\}$ at
	time bin $t \in \{ 1,\ldots,T\}$ as
	$y_{it} \in \mathbb{Z}_{\geq 0}$, and let
	$\bm{y}_{i} =  (y_{i1},\ldots,y_{iT})'$.
	Further, let $z_{i} = j$ denote the cluster indicator of neuron $i$. To facilitate clustering, we re-parametrize the regular Poisson linear dynamical system (PLDS) model to separate the mean log-firing-rate out. Assume neurons are independently Poisson distributed, conditional on the
	low-dimensional latent state $\bm{x}_{t}^{(j)} \in \mathbb{R}^{p}$ as follows:
	\begin{align*}
		y_{it} &\sim Poi(\lambda_{it})\\
		\log\lambda_{it} &= \mu_t^{(j)} + \bm{c}'_i\bm{x}^{(j)}_t
	\end{align*}
	, with $\bm{c}_i\sim N(\bm{0},\bm{I}_p)$. We further assume the intercept $\mu_t^{(j)}$ and the latent state $\bm{x}^{(j)}_t$ progress linearly with Gaussian noise as
	\begin{align*}
		\mu_1^{(j)} &\sim N(\mu_0, \Sigma_0)\\
		\mu_{t+1}^{(j)} &\sim N(f^{(j)}\mu_t^{(j)} + g^{(j)}, \Sigma^{(j)})\\
		\bm{x}_1^{(j)} &\sim N(\bm{x}_0, \bm{Q}_0)\\
		\bm{x}_{t+1}^{(j)} &\sim N(\bm{A}^{(j)}\bm{x}_t^{(j)} + \bm{b}^{(j)}, \bm{Q}^{(j)})
	\end{align*}
	If we denote $\bm{\lambda}_{i} =  (\lambda_{i1},\ldots,\lambda_{iT})'$, $\bm{\mu}^{(j)} = (\mu^{(j)}_1,\ldots,\mu^{(j)}_T)'$ and $\bm{X}^{(j)} = (\bm{x}^{(j)}_1,\ldots,\bm{x}^{(j)}_T)'$, the model can be equivalently written as regular Poisson factor model as
	\begin{align*}
		\bm{y}_i &\sim Poi(\bm{\lambda}_i)\\
		\log\bm{\lambda}_i &= \bm{\mu}^{(j)}+ \bm{X}^{(j)}\bm{c}_i 
	\end{align*}

	However, since the condition $T/N \rightarrow 0$ doesn’t hold \citep{bibid}, the latent state $\bm{X}^{(j)}$ cannot be consistently estimated, and assuming linear dynamics of $\bm{X}^{(j)}$ resolves the problem. Note that when $p>1$, the model is not unique, since $\widetilde{\bm{X}}^{(j)} = \bm{X}^{(j)}\bm{U}$ also satisfies the equation for any orthogonal matrix $\bm{U}$ of order $p$. To ensure the model indefinability, we simply assume $\bm{A}^{(j)}$ and $\bm{Q}^{(j)}$
	are both diagonal for convenience. See more detailed discussions of the constraints in discussion. Overall, denote the cluster-related parameters of cluster $j$ as $\bm{\theta}^{(j)}= \{\bm{\mu}^{(j)}, \bm{X}^{(j)}, f^{(j)}, g^{(j)}, \Sigma^{(j)},\{\bm{A}^{(j)}, \{\bm{b}^{(j)}, \{\bm{Q}^{(j)}\}$ and the spike counts of neuron i is generated by Poisson dynamic factor model (PDFM) as $\bm{Y}_i\sim PDFM(\bm{\theta}^{(z_i)})$, with the prior of $\bm{\theta}^{(j)}$ as $\bm{H}$. The priors for $\{f^{(j)}, g^{(j)}, \Sigma^{(j)},\{\bm{A}^{(j)}, \{\bm{b}^{(j)}, \{\bm{Q}^{(j)}\}$ are regular  normal and inverse-gamma distribution (or multivariate normal and inverse-Wishart when using other non-diagonal constraints).

@@

>>>>

\DeclareMathOperator*{\argmax}{arg\,max}
\DeclareMathOperator*{\argmin}{arg\,min}

@@

	\subsection{Poisson Dynmic Factor Model}
	Denote the observed spike-count of neuron $i \in \{ 1,\ldots,N\}$ at
	time bin $t \in \{ 1,\ldots,T\}$ as
	$y_{it} \in \mathbb{Z}_{\geq 0}$, and let
	$\bm{y}_{i} =  (y_{i1},\ldots,y_{iT})'$.
	Further, let $z_{i} = j$ denote the cluster indicator of neuron $i$. To facilitate clustering, we re-parametrize the regular Poisson linear dynamical system (PLDS) model \citep{Macke2011} to separate the mean log-firing-rate out. Assume neurons are independently Poisson distributed, conditional on the
	low-dimensional latent state $\bm{x}_{t}^{(j)} \in \mathbb{R}^{p}$ as follows:
	\begin{align*}
		y_{it} &\sim Poi(\lambda_{it})\\
		\log\lambda_{it} &= \mu_t^{(j)} + \bm{c}'_i\bm{x}^{(j)}_t
	\end{align*}
	, with $\bm{c}_i\sim N(\bm{0},\bm{I}_p)$. We further assume the intercept $\mu_t^{(j)}$ and the latent state $\bm{x}^{(j)}_t$ progress linearly with Gaussian noise as
	\begin{align*}
		\mu_1^{(j)} &\sim N(\mu_0, \Sigma_0)\\
		\mu_{t+1}^{(j)} &\sim N(f^{(j)}\mu_t^{(j)} + g^{(j)}, \sigma^{2(j)})\\
		\bm{x}_1^{(j)} &\sim N(\bm{x}_0, \bm{Q}_0)\\
		\bm{x}_{t+1}^{(j)} &\sim N(\bm{A}^{(j)}\bm{x}_t^{(j)} + \bm{b}^{(j)}, \bm{Q}^{(j)})
	\end{align*}

	The bias term ($\bm{b}^{(j)}$) in the latent state is constant across time, unlike ind PLDS, which is  $\bm{b}_t^{(j)}$. Allowing the time-varying bias term will lead to infinite solutions of linear dynamics. If the $\bm{A}^{(j)}$ is one solution, then $\bm{A}^{*(j)} = \bm{A}^{(j)} + \bm{M}$ and $\bm{b}_t^{*(j)} = \bm{b}_t^{(j)} - \bm{M}\bm{x}^{(j)}_t$ will be another set of solution, for any $\bm{M} \in \mathbb{R}^{p\times p}$. The same reason for using time constant $g^{(j)}$.

	If we denote $\bm{\lambda}_{i} =  (\lambda_{i1},\ldots,\lambda_{iT})'$, $\bm{\mu}^{(j)} = (\mu^{(j)}_1,\ldots,\mu^{(j)}_T)'$ and $\bm{X}^{(j)} = (\bm{x}^{(j)}_1,\ldots,\bm{x}^{(j)}_T)'$, the model can be equivalently written as regular Poisson factor model as
	\begin{align*}
		\bm{y}_i &\sim Poi(\bm{\lambda}_i)\\
		\log\bm{\lambda}_i &= \bm{\mu}^{(j)}+ \bm{X}^{(j)}\bm{c}_i 
	\end{align*}

	However, since the condition $T/N \rightarrow 0$ doesn’t hold \citep{Johnstone2009}, the latent state $\bm{X}^{(j)}$ cannot be consistently estimated, and assuming linear dynamics of $\bm{X}^{(j)}$ resolves the problem. Note that when $p>1$, the model is not unique, since $\bm{X}^{*(j)} = \bm{X}^{(j)}\bm{U}$ also satisfies the equation for any orthogonal matrix $\bm{U}$ of order $p$. To ensure the model indefinability, we simply assume $\bm{A}^{(j)}$ and $\bm{Q}^{(j)}$
	are both diagonal for convenience. See more detailed discussions of the constraints in discussion. Overall, denote the cluster-related parameters of cluster $j$ as $\bm{\theta}^{(j)}= \{\bm{\mu}^{(j)}, \bm{X}^{(j)}, f^{(j)}, g^{(j)}, \sigma^{2(j)},\bm{A}^{(j)}, \bm{b}^{(j)}, \bm{Q}^{(j)}\}$ and the spike counts of neuron i is generated by Poisson dynamic factor model (PDFM) as $\bm{Y}_i\sim PDFM(\bm{\theta}^{(z_i)})$, with the prior of $\bm{\theta}^{(j)}$ as $\bm{H}$. The priors for $\{f^{(j)}, g^{(j)}, \sigma^{2(j)},\bm{A}^{(j)}, \bm{b}^{(j)}, \bm{Q}^{(j)}\}$ are regular  normal and inverse-gamma distribution (or multivariate normal and inverse-Wishart when using other non-diagonal constraints).

@@

	\subsection{Full Conditional Distributions}

	\underline{\textit{$x_t^{(j)}$ and $\mu_t^{(j)}$}}:
	Let the $t^{\text{th}}$ column observation and firing rate of the $j^{\text{th}}$ cluster be $\widetilde{\bm{y}}_t = vec\left(\{y_{it}^{(j)}|z_i = j\}\right)$ and $\widetilde{\bm{\lambda}}_t^{(j)} = vec\left(\{\lambda_{it}|z_i = j\}\right)$. The number of neurons in cluster $j$ is $n_j=\left|\{i:z_i = j\}\right|$. The corresponding loading for these $n_j$ neurons is $\bm{C}^{(j)}\in\mathbb{R}^{n_j\times p}$, such that $\log \widetilde{\bm{\lambda}}_t^{(j)} = \mu_t^{(j)}\bm{1}_{n_j} + \bm{C}^{(j)}\bm{x}_t^{(j)} = \left(\bm{1}_{n_j}, \bm{C}^{(j)} \right)\cdot \left(\mu_t^{(j)}, \bm{x}_t^{'(j)}\right)'$. Denote $\bm{\widetilde{C}}^{(j)} = \left(\bm{1}_{n_j}, \bm{C}^{(j)} \right)$, $\bm{\widetilde{x}}_t^{(j)} = \left(\mu_t^{(j)}, \bm{x}_t^{'(j)}\right)'$, $\bm{\widetilde{x}}^{(j)} = \left(\bm{\widetilde{x}}_1^{'(j)},\ldots,\bm{\widetilde{x}}_T^{'(j)}\right)'$, $\widetilde{\bm{A}}^{(j)} = diag(f^{(j)}, \bm{A}^{(j)})$, $\widetilde{\bm{b}}^{(j)} = \left(g^{(j)}, \bm{b}^{'(j)}\right)'$ and $\widetilde{\bm{Q}}^{(j)} = diag(\sigma^{2(j)}, \bm{Q}^{(j)})$. The full conditional distribution $P(\bm{\widetilde{x}}^{(j)}|\ldots) = P(\bm{\widetilde{x}}^{(j)}|\{\widetilde{\bm{y}}_t\}_{t=1}^T, \bm{\widetilde{C}}^{(j)}, \widetilde{\bm{A}}^{(j)}, \widetilde{\bm{b}}^{(j)}, \widetilde{\bm{Q}}^{(j)})$ is approximated by a global Laplace approximation, i.e. $P(\bm{\widetilde{x}}^{(j)}|\ldots) \approx  N_{(p+1)T}(\bm{\widetilde{x}}^{(j)} | \bm{\mu}_{\widetilde{\bm{x}}^{(j)}}, \bm{\Sigma}_{\widetilde{\bm{x}}^{(j)}})$, with $\bm{\mu}_{\widetilde{\bm{x}}^{(j)}} = \argmax_{\widetilde{\bm{x}}^{(j)}}P(\bm{\widetilde{x}}^{(j)}|\ldots)$ and $\bm{\Sigma}_{\widetilde{\bm{x}}^{(j)}}) = -\left(\nabla\nabla \log P(\bm{\widetilde{x}}^{(j)}|\ldots)\right |_{\bm{\widetilde{x}}^{(j)} = \bm{\mu}_{\widetilde{\bm{x}}^{(j)}}} )^{-1}$. The log full conditional distribution $h(\bm{\widetilde{x}}^{(j)}) = \log P(\bm{\widetilde{x}}^{(j)}|\ldots)$ is given by:
	\begin{align*}
		h &= \text{const} + \sum_{t=1}^{T} \widetilde{\bm{y}}'_t\bm{\widetilde{C}}^{(j)}\bm{\widetilde{x}}_t^{(j)} - \widetilde{\bm{\lambda}}_t
		- \frac{1}{2}(\bm{\widetilde{x}}_1^{(j)}  - \bm{\widetilde{x}}_0^{(j)})'\bm{\widetilde{\bm{Q}}}^{(j)}_0(\bm{\widetilde{x}}_1^{(j)}  - \bm{\widetilde{x}}_0^{(j)})\\ 
		&-\sum_{t=2}^{T}\frac{1}{2}(\bm{\widetilde{x}}_t^{(j)}  - \widetilde{\bm{A}}^{(j)}\bm{\widetilde{x}}_{t-1}^{(j)} - \widetilde{\bm{b}}^{(j)})'\bm{\widetilde{\bm{Q}}}^{(j)}(\bm{\widetilde{x}}_t^{(j)}  - \widetilde{\bm{A}}^{(j)}\bm{\widetilde{x}}_{t-1}^{(j)} - \widetilde{\bm{b}}^{(j)})
	\end{align*}

	The $h(\bm{\widetilde{x}}^{(j)})$ is concave and hence unimodal, and the Markovian structure of the latent dynamics, and hence the tri-block-diagonal Hessian, makes it possible to compute a Newton update in $\mathcal{O}(T)$ \citep{Paninski2010}.

	To facilitate convergence, we use a smoothing estimate with local Gaussian approximation as a "warm start". The forward filtering for a dynamic Poisson model has been previously described in \cite{Eden2004}. Because the approximated filtering estimates are Gaussian distributed, we can further find the smoothing estimates using a backward pass as in \cite{RAUCH1965}.

	\underline{\textit{$\bm{c}_i$}}: 
	When writing the observation mapping in matrix form,
	$\log\bm{\lambda}_i = \bm{\mu}^{(j)}+ \bm{X}^{(j)}\bm{c}_i$, the update of $\bm{c}_i$ reduces to a regular Poisson regression problem, given $\bm{\mu}^{(j)}$ and $\bm{X}^{(j)}$ are known. Here, we sample the posterior by no a No U-Turn Sampler (NUTS, \cite{Hoffman2011}).

	\underline{\textit{Linear dynamics of latent state}}: The parameters for linear dynamics are $f^{(j)}$, $g^{(j)}$, $\sigma^{2(j)}$, $\bm{A}^{(j)}$, $\bm{b}^{(j)}$ and $\bm{Q}^{(j)}$. To make the model identifiable, we simply assume $\bm{A}^{(j)} = diag(a_1^{(j)},\ldots, a_p^{(j)})$ and $\bm{Q}^{(j)} = diag(q_1^{(j)},\ldots, q_p^{(j)})$. Therefore, we can update $\bm{A}^{(j)}$, $\bm{b}^{(j)}$ and $\bm{Q}^{(j)}$ dimension-by-dimension, as the update in $f^{(j)}$, $g^{(j)}$ and $\sigma^{2(j)}$. Use the priors $\sigma^{2(j)} \sim IG\left(\nu_0/2, \nu_0\sigma_0^2/2\right)$ and  $\left( g^{(j)}, f^{(j)}\right)'\sim N(\bm{\mu}_0, \sigma^{2(j)}\bm{\Lambda}_0^{-1})$. The prior $\mu_0$ is specified as $(0,1)'$. Denote $\bm{y}_{\mu^{(j)}} = \left(\mu_2^{(j)}, \ldots, \mu_T^{(j)}\right)'$ and $\bm{X}_{\mu^{(j)}} = \left(\bm{1}_{T-1}, \bm{\mu}_{1:(T-1)}^{(j)}\right)$, with $\bm{\mu}_{1:(T-1)}^{(j)} = \left(\mu_1^{(j)}, \ldots, \mu_{T-1}^{(j)}\right)'$. The full conditional distributions are $\sigma^{2(j)}|\ldots \sim IG\left(\frac{\nu_0 + T-1}{2}, \frac{\nu_0\sigma^2_0 + \bm{y}'_{\mu^{(j)}}\bm{y}_{\mu^{(j)}} + \bm{\mu}'_0\bm{\Lambda}_0\bm{\mu}_0 - \bm{\mu}'_n\bm{\Lambda}_n\bm{\mu}_n }{2}\right)$ and $\left( g^{(j)}, f^{(j)}\right)'|\ldots \sim N(\bm{\mu}_n, \sigma^{2(j)}\bm{\Lambda}_n^{-1})$, with $\bm{\Lambda}_n = \bm{X}'_{\mu^{(j)}}\bm{X}_{\mu^{(j)}} + \bm{\Lambda}_0$ and $\bm{\mu}_n = \bm{\Lambda}_n^{-1}\left(\bm{X}'_{\mu^{(j)}}\bm{X}_{\mu^{(j)}}\left(\bm{X}'_{\mu^{(j)}}\bm{X}_{\mu^{(j)}}\right)^{-1}\bm{X}'_{\mu^{(j)}}\bm{y}_{\mu^{(j)}} + \bm{\Lambda}_0\bm{\mu}_0\right)$. The updates of $\bm{A}^{(j)}$, $\bm{b}^{(j)}$ and $\bm{Q}^{(j)}$ are similar, with diagonal assumption.
